Implementation of Concurrent Hash Tables in Java

### 2022-04-29 paper.tex
Update paper.tex

\section{Algorithms}
\subsection{Basic Hashtable}
\begin{lstlisting}[language={Java},caption=Source code for the course-grain put() method.,captionpos=b,breaklines=true,frame=single]
public synchronized boolean put(int item) {
    // If capacity if full, do not add
    if (capacity >= ARR_SIZE)
        return false;

@@

    // Check if first try addition
    if (table[key] == null) {
        table[key] = item;
        cleans[key] = false;
    }
    // Otherwise, quadratically probe for empty indices
    else {
        int i = 1;
        int new_key = (key + i) % ARR_SIZE;
        while (table[new_key] != null) {
            i++;
            new_key = (new_key + i) % ARR_SIZE;
        }
        table[new_key] = item;
        cleans[new_key] = false;
    }

@@

The code outlined in "Listing 1" is the put() method for the coarse-grain hash table created for this project. As it is a course grain hash table,
it utilizes the \emph{synchronized} keyword to make a lock on this function, meaning that if one thread were to be in the middle of a call to this function
then any thread that attempted to call it would have to wait until the original thread finished it's function call. Since this implementation uses a fixed size 
hash table, the method will first check to see if the hash table has reached it's maximum capacity. If the hash table is full, then the method returns false,
indicating that the attempt to put the item into the hash table was unsuccessful. In the case that the hash table still has unused memory, then the method
will create a key for the item by hashing it, and checks to see if the memory location associated with that key has been used yet.
\section{Experimental Results}
We began by testing how the standard hash table will performed when run sequentially, and then when multithreaded. 
After those tests, we created a locking hash table and proceeded to run tests on that version. 
After both of those tests, we deicded it would be a good idea to test our hash tables to Java's native concurrent hash map, the closest thing to a concurrent hash table Java has to offer natively.
Our measurement for testing was timing how long each hash table takes to add a specific amount of numbers, make sure each one was added, remove all those numbers, and then make sure each was removed. 
We recorded all the measurements into a CSV file and used a Python program to analyze the numbers, using Pandas to do the math and Matplotlib and Seaborn to visualize the data gathered. 
We were able to gather a plentiful amount of information regarding each hash table and the time it takes to do these processes. 
The data we gathered includes total time taken, the CPU used, the amount of tests, the time it took to do each test for each hash table, the average, low, and high times, and the standard deviation. All times recorded are in millisecconds. 
\begin{figure}[htbp]
    \centerline{\includegraphics[width=90mm,scale=1]{250trials.png}}
    \caption{Graphs of each test for each hash table. The x-axis is the test number, the y-axis is the time in ms the test took. }
    \label{fig}
\end{figure}

>>>>

\section{Algorithms}
\subsection{Coarse-Grain Locking Hashtable}
\begin{lstlisting}[language={Java},caption=Source code for the course-grain put() method.,captionpos=b,breaklines=true,frame=single]
public synchronized boolean put(int item) {
    // If capacity if full, do not add
    if (capacity >= ARR_SIZE)
        return false;

@@

    // Check if first try addition
    if (table[key] == null) {
        table[key] = item;
        cleans[key] = false;
    }
    // Otherwise, linearly probe for empty indices
    else {
        int i = 1;
        int new_key = (key + i) % ARR_SIZE;
        while (table[new_key] != null) {
            i++;
            new_key = (new_key + i) % ARR_SIZE;
        }
        table[new_key] = item;
        cleans[new_key] = false;
    }

@@

The code outlined in "Listing 1" is the \verb|put()| method for the coarse-grain hash table created for this project. As it is a course grain hash table,
it utilizes the \verb|synchronized| keyword to make a lock on this method, meaning that if one thread were to be in the middle of a call to this method
then any thread that attempted to call it would have to wait until the original thread finished it's method call. Since this implementation uses a fixed size hash table, 
the method will first check to see if the hash table has reached it's maximum capacity. If the hash table is full, then the method returns false,
indicating that the attempt to put the item into the hash table was unsuccessful. In the case that the hash table still has unused memory, then the method
will create a key for the item by hashing it, and checks to see if the memory location associated with that key has been used yet. If no other key is associated
with that memory location, then we add the item to the array at that location, and change the clean array so the hash table will know that data was stored 
at this memory location for the linear probing in the \verb|remove()| method. If that memory location already has data associated with it, then the method will
begin linear probing. The way linear probing works is by simply incrementing the key value by one, and checking that memory location. If the memory location
is empty, then the method will store the item at that key value, otherwise, the method will continue to increment and check until it finds a memory location
that is unused. After the value has been put in to the hash table, the hash table's capacity variable is increased to signify how many items are in it, and 
the method call will finally return true to indicate that the value has been added successfully.

\begin{lstlisting}[language={Java},caption=Source code for the course-grain search() method.,captionpos=b,breaklines=true,frame=single]
public synchronized boolean search(int item) {
    int key = hash(item);
    int i = 0;
    int new_key = (key + i) % ARR_SIZE;
    // Quadratically probe for the item
    // Dirty indices are considered to be a collision, the item may of been probed
    // farther down
    while (!cleans[new_key]) {
        if (table[new_key] == null) {
        } else if (item == table[new_key]) {
            return true;
        }
        i++;
        new_key = (new_key + i) % ARR_SIZE;
        if (i >= ARR_SIZE)
            break;
    }

    return false;
}
\end{lstlisting}

The code outlined in "Listing 2" shows the \verb|search()| method for the coarse-grain hash table created for this project. This method's purpose is to accept 
a value, and then check the locations in the hashtable where it could be stored for it. It does this by first acquiring the first key for that item with the
\verb|hash()| method. After the method has the initial key associated with that value, it will check to see if that key is stored in the array at that location.
It does this by creating a loop that checks to see the \verb|clean| array contains any data. What this accomplishes is a simple test, if there is no data in the \verb|clean| array
then the item has never been put into the hash table, otherwise, there is a possibility that the value has been put in and is located here or further into
the hash table due to linear probing. In the case that the locations checked in the \verb|clean| array are ever empty, the while loop is broken and the method returns
false, indicating that the item has not been found. The while loop will continually check different locations inside the \verb|clean| array to see if they are equal to
the item that is being searched for. If the item is found, the method returns true, otherwise the method will loop until every single memory location in the hash table has
been checked and returns false.

\begin{lstlisting}[language={Java},caption=Source code for the course-grain remove() method.,captionpos=b,breaklines=true,frame=single]
public synchronized boolean remove(int item) {
    int key = hash(item);
    int i = 0;
    int new_key = (key + i) % ARR_SIZE;

    // If item not in hash table, no need to look for it again
    if (!search(item)) {
        return false;
    }

    // Quadratically probe for the item
    // Dirty indices are considered to be a collision, the item may of been probed
    // farther down
    while (!cleans[new_key]) {
        if (item == table[new_key]) {
            table[new_key] = null;
            break;
        }
        i++;
        new_key = (new_key + i) % ARR_SIZE;
    }

    // Decrease capacity
    capacity--;
    return true;
}
\end{lstlisting}

\section{Experimental Results}
We began by testing how the standard hash table will performed when run sequentially, and then when multithreaded. 
After those tests, we created a locking hash table and proceeded to run tests on that version. 
After both of those tests, we deicded it would be a good idea to test our hash tables to Java's native concurrent hash map, the closest thing to a concurrent hash table Java has to offer natively.
Our measurement for testing was timing how long each hash table takes to add a specific amount of numbers, make sure each one was added, remove all those numbers, and then make sure each was removed. 
We recorded all the measurements into a CSV file and used a Python program to analyze the numbers, using Pandas to do the math and Matplotlib and Seaborn to visualize the data gathered. 
We were able to gather a plentiful amount of information regarding each hash table and the time it takes to do these processes. 
The data we gathered includes total time taken, the CPU used, the amount of tests, the time it took to do each test for each hash table, the average, low, and high times, and the standard deviation. All times recorded are in millisecconds. 
\begin{figure}[htbp]
    \centerline{\includegraphics[width=90mm,scale=1]{250trials.png}}
    \caption{Graphs of each test for each hash table. The x-axis is the test number, the y-axis is the time in ms the test took. }
    \label{fig}
\end{figure}
